Union and intersection contracts: theory vs practice

### 2021-03-23 paper.tex
Improve contract part

This paper is centered around the design of the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program eventually evaluates to a value that can be
directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
easy to write and easy to embed in a larger system, to serve either directly as
a native language to specify configuration, or as a meta-language to generate
configuration. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages ever built\cite{repology}.
Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well.  While there's no hesitation in incorporating recent research
when it serves a well-defined purpose, the design decisions are guided by the
industrial use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: if your program is only run once on
fixed inputs and terminate, as are such programmable configurations, any type
error will be reported at run-time anyway. Why bother with the complexity of a
static type system? On the other hand, more and more software systems offload
complexity to configurations, such as cloud infrastructure, build system, or
operating systems, leading the corresponding configurations to become
increasingly complex, to the point where static types are beneficial. For
reusable code — that is, library functions — static types brings all the usual,
widely acknowledged benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}, an
approach capable of reuniting static typing and dynamic typing. Unlike
gradually-typed language like TypeScript, we do not seek to build a complex type
system that tries hard to type most of the idioms naturally arising in
dynamically typed code.  Rather, we choose to provide a reasonably expressive
type system with good inference properties, that should make writing typed
generic function that operate on data relatively easy, but may require resorting
to untyped code for more exotic expressions.  However, this is not a surrender,
as Nickel provides a complementary mechanism for more complex data validation,
contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time predicate checking that
nicely supports higher-order functions. It turns out these contracts are closely
related to the type casts introduced by gradual typing. In
\cite{WellTypedBlamed}, the authors precisely propose a system mixing gradual
typing and contracts \textit{à la Findler \& Felleisen}. Nickel adopts a similar
type system, with both statically typed terms, dynamically typed terms, and
first-class contracts. In this setting, each construct of the type system -
arrow types, record types, polymorphic types - must have a contract counterpart,
that is able to check at run-time the membership of a value to the corresponding
type.

>>>>

The explorative work presented in this paper is motived by the design of the
Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that can be
directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
easy to write and easy to embed in a larger system, to serve either directly as
a native language to specify configuration, or as a meta-language to generate
configuration. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages ever built\cite{repology}.  Nickel targets
build systems and cloud deployment tools (Kubernetes, Terraform or NixOps) as
well. While there's no hesitation in incorporating recent research when it
serves a well-defined purpose, the design decisions are guided by the industrial
use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: since your program is run on fixed
inputs and is expected to terminate, any relevant type error will be reported
anyway.  Why bother with the complexity of a static type system? On the other
hand, more and more software systems offload complexity to configurations, such
as cloud infrastructure, build system, or operating systems, leading the
corresponding configurations to become increasingly complex, to the point where
static types are beneficial. For reusable code — that is, library functions —
static types brings all the usual, widely acknowledged benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
that mixes both static typing and dynamic typing. Unlike gradually-typed
language like TypeScript, we do not seek to build a complex type system that
tries hard to make most of the idioms naturally arising in dynamically typed
code typable. Rather, we choose to provide a reasonably expressive type system
with good inference properties, that make usual generic function operating on
data easily typable, but may require resorting to untyped code for more exotic
expressions.  However, this is not a complete surrender, as Nickel provides a
complementary mechanism for more advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is fundamental for good error reporting. It turns out these contracts are
closely related to the type casts introduced by gradual typing, modulo blame:
both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of
contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely a system integrating gradual
typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts a
similar type system, with both statically typed terms, dynamically typed terms,
and first-class contracts. In this setting, each type constructor of the static
type system - arrows, records, foralls - must have a contract counterpart that
is able to check at run-time that a value is indeed a member of the
corresponding type.
